---
title: "Smart_Devices"
author: "Ahmed"
date: '2022-05-30'
output: html_document
---

## Setting up environment
### Installing and loading required packages to set up my R environment

```{r}
install.packages("tidyverse")
install.packages("dplyr")
install.packages("readr")
install.packages("tidyr")
install.packages("lubridate")
install.packages("skimr")
install.packages("janitor")
install.packages("ggplot2")
```

```{r}
library(tidyverse)
library(dplyr)
library(readr)
library(tidyr)
library(lubridate)
library(skimr)
library(janitor)
library(ggplot2)
```


## Importing datasets
### Loading the datasets I will be using for the analysis

Note: Datasets were downloaded from [here](https://www.kaggle.com/datasets/arashnic/fitbit) and were originally sourced from Amazon Mechanical Turk between 03.12.2016-05.12.2016.
```{r}
Daily_Intensities <- read_csv("Daily_Intensities.csv")
Daily_Steps <- read_csv("Daily_Steps.csv")
```


## Observe datasets
### Looking at the datasets and the structures

```{r}
head(Daily_Intensities)
head(Daily_Steps)
```

```{r}
colnames(Daily_Intensities)
str(Daily_Intensities)
skim_without_charts(Daily_Intensities)
```

```{r}
colnames(Daily_Steps)
str(Daily_Steps)
skim_without_charts(Daily_Steps)
```


## Cleaning the datasets
### Creating new data frames with the clean data I will be using

Note: Created new data frames using only data we need for our analysis and formatted date structure for activity levels
```{r}
Daily_Activity <- Daily_Intensities %>% 
  select(Id, ActivityDay, FairlyActiveMinutes, VeryActiveMinutes) %>% 
  mutate(TotalActiveMinutes = FairlyActiveMinutes + VeryActiveMinutes) %>%
  mutate(ActivityDay = mdy(ActivityDay))

glimpse(Daily_Activity)
```

Note: Formatted date structure for step counts
```{r}
Daily_Steps <- Daily_Steps %>% 
  mutate(ActivityDay = mdy(ActivityDay))

glimpse(Daily_Steps)
```


## Analyzing the data
### Creating new data frames to analyze the data and observe trends by daily average

Note: Checking for outliers and inconsistent data
```{r}
sort(table(Daily_Activity$Id))
sort(table(Daily_Steps$Id))
```

Note: Created new data frames showing averages for each user, only taking data from users with enough data to be consistent
```{r}
Average_Activity <- Daily_Activity %>% 
  group_by(Id) %>% 
  filter(n()>=18) %>% 
  summarise(AvgDailyActivity = mean(TotalActiveMinutes))

Average_Steps <- Daily_Steps %>% 
  group_by(Id) %>% 
  filter(n()>=18) %>% 
  summarise(AvgDailySteps = mean(StepTotal))
```

Note: Created new data frame to combine activity data and step count data by daily average
```{r}
Total_Daily_Averages <- Average_Activity %>% 
  inner_join(Average_Steps, by = "Id")
```

```{r}
head(Total_Daily_Averages)
glimpse(Total_Daily_Averages)
```


## Analyzing the data (cont.)
### Creating new data frames to analyze the data and observe trends by type of day average

Note: Checking type of day structure, weekend vs. weekday
```{r}
table(wday(Daily_Activity$ActivityDay, label = FALSE))
table(wday(Daily_Activity$ActivityDay, label = TRUE))
```

Note: Created data frames for average activity based on type of day
```{r}
Weekday_Activity <- Daily_Activity %>% 
  group_by(Id) %>% 
  filter(n()>=18) %>%
  filter(wday(ActivityDay) != 1) %>% 
  filter(wday(ActivityDay) != 7) %>% 
  summarise(AvgWeekdayActivity = mean(TotalActiveMinutes))

Weekend_Activity <- Daily_Activity %>% 
  group_by(Id) %>% 
  filter(n()>=18) %>%
  filter(wday(ActivityDay) != 2) %>% 
  filter(wday(ActivityDay) != 3) %>% 
  filter(wday(ActivityDay) != 4) %>% 
  filter(wday(ActivityDay) != 5) %>% 
  filter(wday(ActivityDay) != 6) %>% 
  summarise(AvgWeekendActivity = mean(TotalActiveMinutes))
```


Note: Created data frames for average step count based on type of day
```{r}
Weekday_Steps <- Daily_Steps %>% 
  group_by(Id) %>% 
  filter(n()>=18) %>%
  filter(wday(ActivityDay) != 1) %>% 
  filter(wday(ActivityDay) != 7) %>% 
  summarise(AvgWeekdaySteps = mean(StepTotal))

Weekend_Steps <- Daily_Steps %>% 
  group_by(Id) %>% 
  filter(n()>=18) %>%
  filter(wday(ActivityDay) != 2) %>% 
  filter(wday(ActivityDay) != 3) %>% 
  filter(wday(ActivityDay) != 4) %>% 
  filter(wday(ActivityDay) != 5) %>% 
  filter(wday(ActivityDay) != 6) %>% 
  summarise(AvgWeekendSteps = mean(StepTotal))
```

Note: Created new data frame to combine activity data and step count data by type of day average
```{r}
Day_Type_Averages <- Weekday_Activity %>% 
  inner_join(Weekend_Activity, by = "Id") %>% 
  inner_join(Weekday_Steps, by = "Id") %>% 
  inner_join(Weekend_Steps, by = "Id")

head(Day_Type_Averages)
glimpse(Day_Type_Averages)
```


## Plotting the data
### Creating a graph to compare the activity and step count data

```{r}
ggplot(data = Total_Daily_Averages) +
  geom_smooth(mapping = aes(x = AvgDailySteps, y = AvgDailyActivity)) +
  geom_point(mapping = aes(x = AvgDailySteps, y = AvgDailyActivity)) +
  labs(title = "Daily Steps compared to Daily Activity", x = "Average Daily Steps", y = "Average Daily Activity (min.)")
```


## Exporting the data
### Creating .csv files to export the data for further analysis and visualization in spreadsheets and Tableau

```{r}
write_csv(Total_Daily_Averages, "Total_Daily_Averages.csv")
write_csv(Day_Type_Averages, "Day_Type_Averages.csv")
```

